Containers and R---the Rockerverse and beyond

### 2019-12-07 manuscript.Rmd
rework section on clients and add overview table

---
title: "Containers and R---the Rockerverse and beyond"
author:
  - name: Daniel Nüst
    affiliation: University of Münster
    address:
      - Institute for Geoinformatics
      - Heisenbergstr. 2
      - 48149 Münster, Germany
      - \href{https://orcid.org/0000-0002-0024-5046}{\includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} https://orcid.org/0000-0002-0024-5046}
    email: daniel.nuest@uni-muenster.de
  - name: Carl Boettiger
    #affiliation: -
    #email: -
  - name: Robrecht Cannoodt
    #affiliatin: -
    #email: -
  - name: Dirk Eddelbuettel
    affiliation: University of Illinois at Urbana-Champaign
    address:
      - Department of Statistics
      - Illini Hall, 725 S Wright St
      - Champaign, IL 61820, USA
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0001-6419-907X}{https://orcid.org/0000-0001-6419-907X}
    email: dirkd@eddelbuettel.com
  - name: Mark Edmondson
    affiliation: IIH Nordic A/S, Google Developer Expert for GCP
    email: mark@markedmondson.me
  - name: Colin Fay
    affiliation: ThinkR
    address:
      - 5O rue Arthur Rimbaud
      - 93300 Aubervilliers, France
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0001-7343-1846}{https://orcid.org/0000-0001-7343-1846}
    email: contact@colinfay.me
  - name: Ben Marwick
    #affiliation: -
    #email: -
  - name: Karthik Ram
    affiliation: Berkeley Institute for Data Science
    address:
      - University of California
      - Berkeley, CA 94720, USA
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0233-1757}{https://orcid.org/0000-0002-0233-1757}
    email: karthik.ram@berkeley.edu
  - name: Noam Ross
    affiliation: EcoHealth Alliance
    address:
      - 460 W 34th St., Ste. 1701
      - New York, NY 10001, USA
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0233-1757}{https://orcid.org/0000-0002-0233-1757}
    email: ross@ecohealthalliance.org
  - name: Nan Xiao
    affiliation: Seven Bridges Genomics
    address:
      - 529 Main St, Suite 6610
      - Charlestown, MA 02129, USA
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0250-5673}{https://orcid.org/0000-0002-0250-5673}
    email: me@nanx.me
  - name: Lori Shepherd
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
      - Elm & Carlton Streets
      - Buffalo, New York, 14263, USA
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-5910-4010}{https://orcid.org/0000-0002-5910-4010}
    email: lori.shepherd@roswellpark.org
  - name: Nitesh Turaga
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
    - Elm & Carlton Streets
    - Buffalo, New York, 14263, USA
    - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0224-9817}{https://orcid.org/0000-0002-0224-9817}
    email: nitesh.turaga@roswellpark.org
abstract: >
  The Rocker project provides widely-used Docker images for R across different application scenarios.
  This articles surveys describe downstream projects building upon Rocker.
  We also look beyond Rocker to other projects connecting containerisation with R.
  These use cases and the diversity of applications demonstrate the power of Rocker and containerisation for collaboration, effectivity, scalability, and transparency.
# blank footnote? https://tex.stackexchange.com/questions/170511/footnotes-without-numbering
output:
  rticles::rjournal_article:
    keep_tex: true
---

@@

## Clients [Ellis Hughes]

Interfacing with the Docker daemon is done through the Docker Command Line Interface (CLI).
Moving back and forth between an R session and the command line can create friction in workflows and reduce reproducibility.
To solve this problem there are a number first order R packages exist that provide an interface to the Docker CLI.
This then means that R users are able to take advantage of the benefits of Docker without explicitly working with the CLI.

- [`stevedore`](https://github.com/richfitz/stevedore)
- [`docker`](https://bhaskarvk.github.io/docker/)
- [`harbor`](https://github.com/wch/harbor/)
- [`dockermachine`](https://github.com/cboettig/dockermachine)
- [`dockyard`](https://github.com/thebioengineer/dockyard)

At the time of writing this paper, these packages can work in concert to provide interfaces for installing Docker, building images, and running/controlling the life cycle of docker containers.

Each of these packages provide functional wrappers to the Docker CLI at different stages of the docker lifecycle.

_dockyard_ has the goal of lowering barrier to creating dockerfiles and docker images, and deploying docker containers. 
The way docker images are made is by reading and following a set of instructions found in dockerfiles/
This requires stepping out of the R ecosystem and into an infrastructure mindset where the user needs to specify the environment they require.
Because dockerfiles can be read as a series of steps, this lends itself well to the `tidyverse` style of programming where each step is piped into the next.
_dockyard_ has a 'dockerfile' object that either starts out as an empty dockerfile, or can use existing dockerfile to use as a template.
This combined with a series of functions that are analogous to the steps commonly found in a dockerfile, it makes building docker images much more approachable to R users.
There are also wrappers for commonly used steps for R programmers like adding libraries and copying files into the R docker containers.

Once the dockerfile is created, _dockyard_ has the functionality to build the image and deploy it.

These clients are important for bringing accessibility to the Docker ecosystem for the wide variety of R programmers.

@@

- [`babelwhale`](https://cran.r-project.org/web/packages/babelwhale/index.html): Running a Docker from R with Singularity or Docker as back-end. This is really useful in HPC environments where a user might not have root access but is able to install Singularity instead of Docker. [@rcannood]
- `RSelenium`
- [`googleComputeEngineR`](https://cloudyr.github.io/googleComputeEngineR/) (function `docker_run`)
  - I don't think `docker_run` is main focus for containers for this package, its more how it uses Docker to launch custom environments that can be templated (e.g. rstudio, shiny etc.) and for parrallisation e.g. `library(future)`  using GCE VMs as endpoints. [@MarkEdmondson1234]
- [`analogsea`](https://github.com/sckott/analogsea) (digital ocean R client)
- [`plumber`](https://www.rplumber.io/docs/hosting.html#docker)
- [production deployment of neural networks](https://github.com/tmobile/r-tensorflow-api) [@jnolis, @nolistic]

>>>>

---
title: "Containers and R---the Rockerverse and beyond"
author:
  - name: Daniel Nüst
    affiliation: University of Münster
    address:
      - Institute for Geoinformatics
      - Heisenbergstr. 2
      - 48149 Münster, Germany
      - \href{https://orcid.org/0000-0002-0024-5046}{\includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} https://orcid.org/0000-0002-0024-5046}
    email: daniel.nuest@uni-muenster.de
  - name: Carl Boettiger
    #affiliation: -
    #email: -
  - name: Robrecht Cannoodt
    affiliation: Ghent University
    address:
      - Data Mining and Modelling for Biomedicine group
      - VIB Center for Inflammation Research
      - Technologiepark 71
      - 9052 Ghent, Belgium
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0003-3641-729X}{https://orcid.org/0000-0003-3641-729X}
    #email: -
  - name: Dirk Eddelbuettel
    affiliation: University of Illinois at Urbana-Champaign
    address:
      - Department of Statistics
      - Illini Hall, 725 S Wright St
      - Champaign, IL 61820, USA
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0001-6419-907X}{https://orcid.org/0000-0001-6419-907X}
    email: dirkd@eddelbuettel.com
  - name: Mark Edmondson
    affiliation: IIH Nordic A/S, Google Developer Expert for GCP
    email: mark@markedmondson.me
  - name: Colin Fay
    affiliation: ThinkR
    address:
      - 5O rue Arthur Rimbaud
      - 93300 Aubervilliers, France
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0001-7343-1846}{https://orcid.org/0000-0001-7343-1846}
    email: contact@colinfay.me
  - name: Ben Marwick
    #affiliation: -
    #email: -
  - name: Karthik Ram
    affiliation: Berkeley Institute for Data Science
    address:
      - University of California
      - Berkeley, CA 94720, USA
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0233-1757}{https://orcid.org/0000-0002-0233-1757}
    email: karthik.ram@berkeley.edu
  - name: Noam Ross
    affiliation: EcoHealth Alliance
    address:
      - 460 W 34th St., Ste. 1701
      - New York, NY 10001, USA
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0233-1757}{https://orcid.org/0000-0002-0233-1757}
    email: ross@ecohealthalliance.org
  - name: Nan Xiao
    affiliation: Seven Bridges Genomics
    address:
      - 529 Main St, Suite 6610
      - Charlestown, MA 02129, USA
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0250-5673}{https://orcid.org/0000-0002-0250-5673}
    email: me@nanx.me
  - name: Lori Shepherd
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
      - Elm & Carlton Streets
      - Buffalo, New York, 14263, USA
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-5910-4010}{https://orcid.org/0000-0002-5910-4010}
    email: lori.shepherd@roswellpark.org
  - name: Nitesh Turaga
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
    - Elm & Carlton Streets
    - Buffalo, New York, 14263, USA
    - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0224-9817}{https://orcid.org/0000-0002-0224-9817}
    email: nitesh.turaga@roswellpark.org
abstract: >
  The Rocker project provides widely-used Docker images for R across different application scenarios.
  This articles surveys describe downstream projects building upon Rocker.
  We also look beyond Rocker to other projects connecting containerisation with R.
  These use cases and the diversity of applications demonstrate the power of Rocker and containerisation for collaboration, effectivity, scalability, and transparency.
# blank footnote? https://tex.stackexchange.com/questions/170511/footnotes-without-numbering
output:
  rticles::rjournal_article:
    keep_tex: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(kableExtra)
```

@@

## Clients [Ellis Hughes, Robrecht Cannoodt @rcannood]

Interfacing with the Docker daemon is typically done through the Docker Command Line Interface (CLI).
However, moving back and forth between an R console and the command line can create friction in workflows and reduce reproducibility.
A number of first-order R packages provide a interface to the Docker CLI, allowing to automate interaction with the Docker CLI from an R console.

Each of these packages has particular advantages as they provide function wrappers for interacting with the Docker CLI at different stages of a container's life cycle. Examples of such interactions are installing the Docker software, creating Dockerfiles, building an image, and launching a container on a local machine or on the cloud. As such, the choice of which package is most useful depends on the use-case at hand.
<!-- ^ in the examples listed above, the name of the packages which provide such functionality could be mentioned, this would highlight the variety in functionality already. -->

```{r clientfunctionalities, echo=FALSE, results='asis', warning=FALSE, message=FALSE,}

library(kableExtra)

functionalities <- tribble(
  ~id, ~Functionality,
  "dockerfile", "Generate a Dockerfile",
  "build", "Build an image",
  "execute", "Execute a container locally or remotely",
  "cloud", "Deploy or manage an instances in the cloud",
  "interact", "Interact with an instance (e.g. file transfer)",
  "images", "Manage storage of images",
  "othercontainers", "Supports Docker and Singularity",
  "api", "Direct access to Docker API instead of using the CLI",
  "install", "Installing Docker software"
)

# some functionalities might be missing. please add them if you find that any are missing!
package_functionalities <- tribble(
  ~package, ~id,
  "stevedore", c("api", "execute", "cloud", "images", "interact"), # build?
  "dockyard", c("dockerfile", "build", "execute"), # cloud?
  "dockermachine", c("install", "execute", "cloud", "interact"),
  "AzureContainers", c("execute", "cloud"),
  "babelwhale", c("execute", "interact", "othercontainers"),  
  "harbor", c("execute", "cloud", "images")
)

tab <- 
  full_join(
  functionalities %>% crossing(package = package_functionalities$package), 
  package_functionalities %>% unnest(id) %>% mutate(check = "\\checkmark"),#\u2714 # \\checkmark
  by = c("id", "package")
) %>%
  mutate(check = ifelse(is.na(check), "", check)) %>% 
  spread(package, check) %>% 
  slice(match(functionalities$id, id)) %>% 
  select(-id)

knitr::kable(tab, escape = FALSE) %>% 
  # kable_styling(latex_options = "striped") %>%
  row_spec(0, angle = -90)
# don't print this table just yet, wait for this PR to be merged
# https://github.com/rstudio/rticles/pull/261
```

**stevedore** ([https://cran.r-project.org/package=stevedore](https://cran.r-project.org/package=stevedore)) 
<!-- from the CRAN description:
Work with containers over the Docker API. Rather than using system calls to interact with a docker client, using the API directly means that we can receive richer information from docker. The interface in the package is automatically generated using the 'OpenAPI' (a.k.a., 'swagger') specification, and all return values are checked in order to make them type stable.
-->

<!--
**docker** ([https://bhaskarvk.github.io/docker](https://bhaskarvk.github.io/docker)
docker has been removed from CRAN, seems unmaintained by authors. Maybe it should be removed from this list?
-->

**dockyard** ([https://github.com/thebioengineer/dockyard](https://github.com/thebioengineer/dockyard)) has the goal of lowering barrier to creating dockerfiles and docker images, and deploying docker containers. 
The way docker images are made is by reading and following a set of instructions found in dockerfiles/
This requires stepping out of the R ecosystem and into an infrastructure mindset where the user needs to specify the environment they require.
Because dockerfiles can be read as a series of steps, this lends itself well to the `tidyverse` style of programming where each step is piped into the next.
dockyard has a 'dockerfile' object that either starts out as an empty dockerfile, or can use existing dockerfile to use as a template.
This combined with a series of functions that are analogous to the steps commonly found in a dockerfile, it makes building docker images much more approachable to R users.
There are also wrappers for commonly used steps for R programmers like adding libraries and copying files into the R docker containers.
Once the dockerfile is created, dockyard has the functionality to build the image and deploy it.

<!-- From the github description:
The goal of **dockermachine** ([https://github.com/cboettig/dockermachine](https://github.com/cboettig/dockermachine)) is to provide a convenient interface to docker-machine from the R command line. This will allow R users to easily create, control, and destroy cloud instances across a wide variety of providers.
-->

<!-- From the cran description
**AzureContainers** ([https://cran.r-project.org/package=AzureContainers](https://cran.r-project.org/package=AzureContainers)) lets you build and deploy containerised services in R, using Docker and Kubernetes. For full functionality, you should have Docker installed, as well as the kubectl and helm commandline tools. Otherwise it is relatively lightweight, requiring neither Powershell nor Python.
-->

**babelwhale** ([https://cran.r-project.org/package=babelwhale](https://cran.r-project.org/package=babelwhale)) 
allows executing and interacting with containers, which can use either Docker or Singularity as a backend.
provides a unified interface to interact with Docker and Singularity containers. You can execute a command inside a container, mount a volume or copy a file.

<!-- from the github description
**harbor** ([https://github.com/wch/harbor](https://github.com/wch/harbor) is an R package for controlling docker containers on local and remote hosts.
-->

@@

- `RSelenium`
- [`googleComputeEngineR`](https://cloudyr.github.io/googleComputeEngineR/) (function `docker_run`)
  - I don't think `docker_run` is main focus for containers for this package, its more how it uses Docker to launch custom environments that can be templated (e.g. rstudio, shiny etc.) and for parrallisation e.g. `library(future)`  using GCE VMs as endpoints. [@MarkEdmondson1234]
- [`analogsea`](https://github.com/sckott/analogsea) (digital ocean R client)
- [`plumber`](https://www.rplumber.io/docs/hosting.html#docker)
- [production deployment of neural networks](https://github.com/tmobile/r-tensorflow-api) [@jnolis, @nolistic]
